Bioinformatics analysis of whole-genome bisulfite-sequencing data

### 2014-12-11 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Removed R chunks

```{r ADS_M-bias, eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "M-bias plot for the ADS sample from the Lister (2011) dataset. Each of read 1 (R1) and read 2 (R2) are plotted separately. We see significant CpG M-bias at the start of read 2 and some noise at the start of read 1.", dev = "pdf", outwidth = "0.8\\linewidth"}
library(dplyr)
library(ggplot2)
thesis_theme <- theme_classic(base_size = 12)
thesis_theme <- theme_bw(base_size = 12)
read.mbias <- function(file) {

  # Echo some information
  message(paste0("Reading ", file, " ..."))

  # Read the file.
  x <- read.table(file, sep = '\n', as.is = TRUE)

  # Find 'header' lines, which are dispersed throughout the file.
  header_lines <- sapply(X = x, function(xx) {
    grepl(pattern = 'context', x = xx)
    })
  y <- vector(mode = "list", length = sum(header_lines))
  names(y) <- x[header_lines]

  # z indexes the lines containing M-bias values.
  z <- sapply(X = x, function(xx) {
    grep(pattern = "context|^=|position", x= xx, invert = TRUE)
    })
  dz <- diff(z)
  next_context_idx <- c(which(dz > 1), nrow(z))
  colnames_idx <- sapply(X = x, function(xx) {
    grep(pattern = "position", x = xx)
    })[1]
  colnames <- strsplit(x = x[colnames_idx, ], split = "\t")[[1]]

  # Loop over the lines in the file.
  for (i in seq_along(next_context_idx)) {
    # Note the "+4" to skip the intermediate 'header' rows
    start_idx <- ifelse(i == 1, 4, z[next_context_idx[i - 1]] + 4)
    end_idx <- z[next_context_idx[i]]  
    yy <- x[seq(from = start_idx, to = end_idx, by = 1), ]
    y[[i]] <- do.call("rbind", lapply(yy, function(yyy, colnames) {
      tmp <- strsplit(yyy, split = "\t")
      val <- data.frame(as.integer(tmp[[1]][1]),
                        as.integer(tmp[[1]][2]),
                        as.integer(tmp[[1]][3]),
                        as.numeric(tmp[[1]][4]),
                        as.integer(tmp[[1]][5]))
      colnames(val) <- colnames
      return(val)
      }, colnames = colnames))
    }
  y <- do.call("rbind", y)
  y$Context <- strtrim(rownames(y), width = 3)

  # Add R1/R2 annotation if paired-end data
  if (grepl(pattern = "R1", x = rownames(y)[1])){
    y$Read <- ifelse(grepl(pattern = "R1", x = rownames(y)), "R1", "R2")
    } else{
      y$Read <- "SE"
      }
  rownames(y) <- NULL

  # Simplify column names
  colnames(y) <- c('read_pos', 'M', 'U', 'perc_M', 'total',
                   'methylation_type', 'read')

  # Re-order columns
  y <- y[, c('read', 'read_pos', 'methylation_type', 'M', 'U', 'total',
             'perc_M')]

  # Replace CpG with CG
  y$methylation_type <- ifelse(y$methylation_type == 'CpG', 'CG',
                               y$methylation_type)

  # Convert to an MBias object.
  return(y)
  }

#files <- paste0("../M-bias/", list.files("../M-bias/"))
# ADS
files <- "../data/M-bias/ADS.M-bias.txt"
sample_names <- gsub("../data/M-bias/", "", gsub(".M-bias.txt", "", files))
mbias <- lapply(files, read.mbias)
mbias <- tbl_df(cbind(sample = rep(sample_names, sapply(mbias, nrow)),
                      do.call("rbind", mbias)))
mbias <- mutate(mbias, read_f = ifelse(read == "SE", "R1", read))
mbias %>%
  filter(sample == "ADS") %>%
  ggplot(aes(x = read_pos, y = perc_M, colour = methylation_type)) +
  geom_line(lwd = 1.3) + facet_grid(. ~ read_f) + ggtitle("M-bias: ADS") +
  ylim(c(0, 100)) + ylab("Percentage methylation") + xlab("Read position") +
  thesis_theme +
  scale_color_discrete(guide = guide_legend(title = "Methylation type"))
```

@@

```{r E18VA_M-bias, eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "M-bias plot for the E18VA sample from the EPISCOPE dataset. Each of read 1 (R1) and read 2 (R2) are plotted separately. For CpGs in read 1, we see noise at the start of the read, followed by a downward slope in the M-bias, which ends with a spike. For CpGs in read 2, we see a downward spike at the start of the read following by a gradual increase in the M-bias curve, with a spike at read-position 101 and a spike at the last read positions for all methylation types. The spike at read-position 101 is also evidence, albeit to a lesser extent, in read 1. This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.", dev = "pdf", outwidth = "0.8\\linewidth"}
#files <- paste0("../M-bias/", list.files("../M-bias/"))
# E18VA
files <- "../data/M-bias/E18VA.M-bias.txt"
sample_names <- gsub("../data/M-bias/", "", gsub(".M-bias.txt", "", files))
mbias <- lapply(files, read.mbias)
mbias <- tbl_df(cbind(sample = rep(sample_names, sapply(mbias, nrow)),
                      do.call("rbind", mbias)))
mbias <- mutate(mbias, read_f = ifelse(read == "SE", "R1", read))
mbias %>%
  filter(sample == "E18VA") %>%
  ggplot(aes(x = read_pos, y = perc_M, colour = methylation_type)) +
  geom_line(lwd = 1.3) + facet_grid(. ~ read_f) + ggtitle("M-bias: E18VA") +
  ylim(c(0, 100)) + ylab("Percentage methylation") + xlab("Read position") +
  thesis_theme +
  scale_color_discrete(guide = guide_legend(title = "Methylation type"))
```

@@

While 1-tuples are the basis of most analyses of bisulfite-sequencing data, these do not make full use of the information available from bisulfite-sequencing data. Each bisulfite-sequencing read may contain multiple methylation loci, which I call an m-tuple. Furthermore, since each read is from a single cell[^chimeric_reads], we can study the methylation calls at m-tuples to learn about _epialleles_ and _epipolymorphisms_ (Chapter \ref{chap:wgbs_statistical_framework}) and _co-methylation_ (Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}). In order to study these concepts, we need software that can perform methylation calling at m-tuples, which was not available when I began my PhD, and so I wrote `methtuple`.

@@

`methtuple` ([https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple)) is software to perform methylation calling at m-tuples from bisulfite-sequencing data. Table \ref{tab:other_m-tuple_methylation_callers} lists other software that can perform methylation calling at m-tuples. However, none of these existed when I started writing `methtuple` nor do any of them do exactly what I require. To the best of my knowledge, `methtuple` is the only software that can perform methylation calling at m-tuples from whole-genome data.

\begin{table}[h]
\centering
\begin{tabulary}{\textwidth}{LLLLL}
\toprule
Software           & Reference                                & Input                                              & Designed for targeted sequencing & Other limitations                                                                \\ \midrule
\texttt{methclone} & \citet{Li:2014ei}                       & Bismark BAM files                                  & No, although example uses eRRBS  & \textbf{TODO: Suspect you can't get actual methylation calls.} \\
\texttt{methpat}  & \url{https://github.com/bjpop/methpat} & Output of \texttt{bismark\_methylation\_extractor} & Yes (amplicons)                  &                                                                                  \\
\texttt{DMEAS}     & \citet{He:2013cj}                       & Output of \texttt{bismark\_methylation\_extractor} &                                  & Windows operating system only. Perl code only available as PDF file.             \\ \bottomrule
\end{tabulary}
\caption{Other software for methylation calling at m-tuples.}
\label{tab:other_m-tuple_methylation_callers}
\end{table}

@@

```{r methtuple_runtime, eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "The running times are the 'User time' reported by \texttt{GNU time} converted from seconds to minutes. The suffix 'ac' on the tuple size means that the option \texttt{--all-combinations} was set.", dev = "pdf", out.width = "0.8\\linewidth"}
library(dplyr)
library(ggplot2)
library(stringr)
library(gridExtra)

# facet_wrap_labeller() from http://stackoverflow.com/a/19298442
facet_wrap_labeller <- function(gg.plot, labels = NULL) {
  #works with R 3.0.1 and ggplot2 0.9.3.1

  g <- ggplotGrob(gg.plot)
  gg <- g$grobs
  strips <- grep("strip_t", names(gg))

  for(ii in seq_along(labels))  {
    modgrob <- getGrob(gg[[strips[ii]]], "strip.text",
                       grep=TRUE, global=TRUE)
    gg[[strips[ii]]]$children[[modgrob$name]] <- editGrob(modgrob,
                                                          label = labels[ii])
    }

  g$grobs <- gg
  class(g) = c("arrange", "ggplot",class(g))
  g
  }
thesis_theme <- theme_classic(base_size = 12)
thesis_theme <- theme_bw(base_size = 12)
# Time
time <- lapply(paste0("../data/methtuple_performance/time/",
                      list.files("../data/methtuple_performance/time/")),
               read.table)
names(time) <- gsub(".time", "",
                    list.files("../data/methtuple_performance/time/"))
n <- str_split(names(time), "\\.")
time <- data_frame("Chromosome" = factor(rep(sapply(n, "[", 1),
                                             times = sapply(time, nrow)),
                                         levels =
                                           paste0("chr",
                                                  c(1:22, "X", "Y", "M", "L"))),
                   "Tuple" = rep(sapply(n, "[", 2),
                                 times = sapply(time, nrow)),
                   "Time" = unlist(time))

gg_time <- time %>%
  ggplot(aes(x = Chromosome, y = Time / 60)) + geom_boxplot() +
  facet_wrap( ~ Tuple, ncol = 3) + thesis_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ylab("Running time (minutes)") +
  ggtitle("Running time of methtuple across 49 samples")
facet_wrap_labeller(gg_time, paste0("Tuple size: ", unique(time$Tuple)))
```

```{r methtuple_memory, eval = TRUE, echo = FALSE, message = FALSE, fig.cap = "The maximum memory usage is the 'Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes. These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}). The suffix 'ac' means that the option \texttt{--all-combinations} was set.", dev = "pdf", out.width = "0.8\\linewidth"}
# Memory
memory <- lapply(paste0("../data/methtuple_performance/memory/",
                        list.files("../data/methtuple_performance/memory/")),
                 read.table)
names(memory) <- gsub(".memory", "",
                      list.files("../data/methtuple_performance/memory/"))
n <- str_split(names(memory), "\\.")
# NB: Need to divide reported memory usage by 4 due to bug in 'time'
# (https://bugzilla.redhat.com/show_bug.cgi?id=703865)
memory <- data_frame("Chromosome" = factor(rep(sapply(n, "[", 1),
                                               times = sapply(memory, nrow)),
                                           levels =
                                             paste0("chr",
                                                    c(1:22, "X", "Y", "M", "L"))),
                     "Tuple" = rep(sapply(n, "[", 2),
                                   times = sapply(memory, nrow)),
                     "Memory" = unlist(memory) / 4)
gg_memory <- memory %>%
  ggplot(aes(x = Chromosome, y = Memory / 10^6)) + geom_boxplot() +
  facet_wrap( ~ Tuple, ncol = 3) + thesis_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ylab("Memory usage (GB)") +
  ggtitle("Maximum memory usage of methtuple across 49 samples")
facet_wrap_labeller(gg_memory, paste0("Tuple size: ", unique(memory$Tuple)))
```

>>>>

\begin{figure}[h!]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/ADS_M-bias.pdf}
\caption{M-bias plot for the ADS sample from the Lister (2011) dataset. Each of read 1 (R1) and read 2 (R2) are plotted separately. We see significant CpG M-bias at the start of read 2 and some noise at the start of read 1.}
\label{fig:ADS_M-bias}
\end{figure}

@@

\begin{figure}[h!]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/E18VA_M-bias.pdf}
\caption{M-bias plot for the E18VA sample from the EPISCOPE dataset. Each of read 1 (R1) and read 2 (R2) are plotted separately. For CpGs in read 1, we see noise at the start of the read, followed by a downward slope in the M-bias, which ends with a spike. For CpGs in read 2, we see a downward spike at the start of the read following by a gradual increase in the M-bias curve, with a spike at read-position 101 and a spike at the last read positions for all methylation types. The spike at read-position 101 is also evidence, albeit to a lesser extent, in read 1. This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.}
\end{figure}

@@

`methtuple` ([https://github.com/PeteHaitch/methtuple](https://github.com/PeteHaitch/methtuple)) is software I wrote to perform methylation calling at m-tuples. In order to explain how it works and my motivation for writing it, I first need to explain what an m-tuple is.

@@

While 1-tuples are the basis of most analyses of bisulfite-sequencing data, these do not make full use of the information available from bisulfite-sequencing data. Each bisulfite-sequencing read may contain multiple methylation loci, which I call an m-tuple. Furthermore, since each read is from a single cell[^chimeric_reads], we can study the methylation calls at m-tuples to learn about _epialleles_ and _epipolymorphisms_ (Chapter \ref{chap:wgbs_statistical_framework}) and _co-methylation_ (Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}).

In order to study these concepts, we need software that can perform methylation calling at m-tuples. Table \ref{tab:other_m-tuple_methylation_callers} lists other software that can perform methylation calling at m-tuples. However, none of these existed when I started writing `methtuple`. Even now, none of these do exactly what I require and have more general limitations. To the best of my knowledge, `methtuple` is the only software that can perform methylation calling at m-tuples from whole-genome data.

\begin{table}[h]
\centering
\begin{tabulary}{\textwidth}{LLLL}
\toprule
Software           & Reference                                & Input                                              & Limitations                                                                \\ \midrule
methclone & \citet{Li:2014ei}                       & \texttt{Bismark} \texttt{BAM}                                  & Unable to install \\
methpat  & \url{https://github.com/bjpop/methpat} & Output of \texttt{bsm} & Designed for amplicons not whole-genome data.                                                                                 \\
DMEAS    & \citet{He:2013cj}                       & Output of \texttt{bsm} & Windows operating system only. Perl code only available as PDF file.             \\
\bottomrule
\end{tabulary}
\caption{Other software for methylation calling at m-tuples and their limitations. Abbreviations: \texttt{bsm} =  \texttt{bismark\_methylation\_extractor}.}
\label{tab:other_m-tuple_methylation_callers}
\end{table}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_runtime.pdf}
\caption{The running times are the `User time' reported by \texttt{GNU time} converted from seconds to minutes. The suffix `ac' on the tuple size means that the option \texttt{--all-combinations} was set. Note that the reported $49$ samples separately counts each of the Ziller sequencing runs.}
\label{fig:methtuple_runtime}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_memory.pdf}
\caption{The maximum memory usage is the `Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes. These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}). The suffix `ac' means that the option \texttt{--all-combinations} was set. Note that the reported $49$ samples separately counts each of the Ziller sequencing runs.}
\label{fig:methtuple_runtime}
\end{figure}
